---
title: "Exporatory data analysis by querying the ToppGene Suite"
shorttitle: "toppgene"
author:
  - name: Pariksheet Nanda
    affiliation: University of Pittsburgh
    email: pan79@pitt.edu
package: toppgene
abstract: >  
  The ToppGene Suite is a one-stop portal for gene list enrichment analysis and
  candidate gene prioritization based on functional annotations and protein
  interactions network.  Although the ToppCluster web application provides
  convenient graphical access to the ToppGene Suite, the OpenAPI 3.0 compliant
  interface of ToppGene is better suited for automation.  This package was
  initial generated from OpenAPI Generator and supplemented with Bioconductor
  class interfaces and more relevant biological examples.
bibliography: references.bib
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{toppgene}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts}
#| include = FALSE
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

<!-- markdownlint-disable MD025 -->

# Overview

The `r BiocStyle::Biocpkg("toppgene")` package
is a client for the ToppGene Suite webserver
that takes as input one or more gene lists
to perform enrichment analysis.

To demonstrate the use of ToppGene,
below are the two test cases from the publication [@chen_improved_2007]
of congenital heart disease (CHD) and diabetic retinopathy (DR).

# Usage

## Prepare the gene lists

A query may contain one or more `GeneSet` objects.
To run a query with multiple `GeneSet` objects,
assemble them into a `GeneSetCollection`.
The published example provides gene symbols for CHD (n = 28) DR and (n = 27).
To use gene identifiers other than gene symbols,
see the `r BiocStyle::Biocpkg("GSEABase")` package.

ToppGene identifies genes by Entrez IDs.
You can convert gene symbols into Entrez IDs
either using Bioconductor or using ToppGene's Lookup feature.

```{r setup}
genes_chd_sym <- c(
    "ADD1", "CITED2", "DTNA", "CKM", "GATA4", "GJA1", "HAND1", "HAND2", "HEY2",
    "HOXC4", "HOXC5", "ITGB3", "JARID2", "MTHFD1", "MTHFR", "MTRR", "NKX2-5",
    "NOS3", "NPPA", "NPPB", "RFC1", "SALL4", "TBX1", "TBX5", "TBX20",
    "TGFB1", "ZFPM2", "ZIC3")
genes_dr_sym <- c(
    "ACE", "ADRB3", "AGT", "AGTR2", "AKR1B1", "APOE", "AR", "CMA1", "EDN1",
    "GNB3", "HFE", "HLA-DPB1", "HLA-DRB1", "ICAM1", "ITGA2B", "ITGB2", "LTA",
    "NOS2A", "NOS3", "NPY", "PECAM1", "PON1", "RAGE", "SELE", "SERPINE1",
    "TIMP3", "TNF")
```

## Convert gene symbol IDs to Entrez IDs

```{r toppgene-convert}
library(toppgene)

genes_chd <- lookup(genes_chd_sym)
genes_chd

genes_dr <- lookup(genes_dr_sym)
genes_dr
```

## Run enrichment query

```{r toppgene-query}
enrich_chd <- enrich(genes_chd$Entrez)
enrich_chd
```

# Session Info {.unnumbered}

```{r session-info}
sessionInfo()
```

# References {.unnumbered}
